---
title: "DLUHC style for R graphics"
date: "Last updated: `r format(Sys.Date())`"
output: 
  github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE, echo=TRUE, warning=FALSE}
pacman::p_load('dplyr', 'tidyr', 'gapminder',
               'ggplot2',  'ggalt',
               'forcats', 'R.utils', 'png', 
               'grid', 'ggpubr', 'scales',
               'dluhctheme', 'knitr', 'pander')

options(scipen = 999)
```

### Credit to BBC data team
This package has been copied from the much better BBC bbplot, and reapplied to work for creating graphics in a DLUHC style. Credit must go to the BBC data team for their fantastic product which the team at DLUHC have essentially copied for our own use.

### Install the dluhctheme package

`dluhctheme` is not on CRAN, so you will have to install it directly from Github using `devtools`. 

If you do not have the `devtools` package installed, you will have to run the first line in the code below as well. 

``` {r eval = FALSE}
# install.packages('devtools')
devtools::install_github('communitiesuk/dluhctheme')
```

For more info on `dluhctheme` check out the [package's Github repo](https://github.com/communitiesuk/dluhctheme), but most of the details about how to use the package and its functions are detailed below. 

When you have downloaded the package and successfully installed it you are good to go and create charts. 

### How does the dluctheme package work?

The package has two types of functions;

* Generic functions:
`dluhc_style()` and `finalise_plot()` are functions which are designed to be generic themes which can be added to any ggplot2 object. They only edit the axes and the grid lines and define the text style, they do not set the colour of lines or fill on the graphs.

* Specific functions:
The other functions are used as a quick tool for instantly creating a particular type of formatted graph based on a set shape of data inputted. 
These are less adaptable, but allow a user to create a graph in one line of code:
`one_line_timeseries()`, `multi_line_timeseries()`, `facet_highlight_timeseries()`, `LA_map()`, `facet_barchart()`, `forecast_timeseries()`

These functions allow you to create a data visualization in a set theme instantly but are more fixed in some of their formatting. Examples of these functions can all be found in the "Reference" tab at the top of this page 
